---
title: "TSSr Vignette"
author: "Zhaolian Lu, Keenan Berry, Zhenbin Hu, Yu Zhan, Tae-Hyuk Ahn, Zhenguo Lin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TSSr Vignette}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

TSSr is a comprehensive R/Bioconductor package for analyzing transcription
start site (TSS) sequencing data. It supports multiple input formats including
BAM, BED, BigWig, and TSS tables, and provides a complete workflow from TSS
identification through downstream analyses such as core promoter shape analysis,
gene annotation, differential expression, and promoter shifting.

## Installation

Install TSSr through Bioconductor with `BiocManager`:

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("TSSr")
```

And load `TSSr`:

```{r library, results='hide', message=FALSE}
library(TSSr)
```

# Citation

If you use TSSr, please cite the following article:

```{r citation, eval=TRUE}
citation("TSSr")
```

# Getting help

For general questions about the usage of TSSr, use the [official Bioconductor support forum](https://support.bioconductor.org) and tag your question "TSSr". We strive to answer questions as quickly as possible.

For technical questions, bug reports and suggestions for new features, we refer to the [TSSr github page](https://github.com/Linlab-slu/TSSr/issues).

# Quick start

This vignette runs an analysis from input files rather than beginning with a
precomputed `TSSr` object. The same object, named `myTSSr`, is returned and
reassigned at every stage of the workflow.

## Bundled input files

The example TSS table contains four samples. The accompanying GFF file
provides gene coordinates for annotation. `system.file()` locates these files
inside the installed package, and `mustWork = TRUE` ensures that a missing file
is reported immediately.

```{r bundled-input-files}
exampleInput <- system.file(
    "extdata",
    "example-tss-table.tsv",
    package = "TSSr",
    mustWork = TRUE
)
exampleAnnotation <- system.file(
    "extdata",
    "example-annotation.gff3",
    package = "TSSr",
    mustWork = TRUE
)
```

We create one `TSSr` object named `myTSSr` and retain this name throughout the
vignette. The four input samples are assigned to two merged groups: `SL01` and
`SL02` form the control group, while `SL03` and `SL04` form the treat group.

```{r construct-object}
myTSSr <- TSSr(
    genomeName = "BSgenome.Scerevisiae.UCSC.sacCer3",
    inputFiles = exampleInput,
    inputFilesType = "TSStable",
    sampleLabels = c("SL01", "SL02", "SL03", "SL04"),
    sampleLabelsMerged = c("control", "treat"),
    mergeIndex = c(1, 1, 2, 2),
    refSource = exampleAnnotation
)

myTSSr
```

## Using your own data

To analyse your own data, replace the values in the `TSSr()` call above while
keeping the object name `myTSSr`. The supplied values must satisfy the
requirements below.

### Constructor arguments

| Argument | Required | Accepted values and restrictions |
|:--|:--:|:--|
| `genomeName` | Yes | A single character string naming an installed `BSgenome` package that matches the genome assembly used to generate the input files, for example `"BSgenome.Scerevisiae.UCSC.sacCer3"`. |
| `inputFiles` | Yes | A character vector containing paths to existing regular files. For `"bam"`, `"bamPairedEnd"`, `"bed"`, `"tss"`, and `"BigWig"`, supply one file per original sample. For `"TSStable"`, supply exactly one table containing all samples. |
| `inputFilesType` | Yes | One case-sensitive value chosen from `"bam"`, `"bamPairedEnd"`, `"bed"`, `"tss"`, `"BigWig"`, or `"TSStable"`. See the table below. |
| `sampleLabels` | Yes | A non-empty character vector naming the original samples. Except for `"TSStable"`, its length must equal `length(inputFiles)`. For `"TSStable"`, the labels identify sample columns in the single input table. |
| `sampleLabelsMerged` | No | A character vector naming the groups produced by replicate merging, in the order indicated by `mergeIndex`. It must be supplied together with `mergeIndex`. If both arguments are omitted, each sample is treated as a separate group. |
| `mergeIndex` | No | A numeric vector assigning every original sample to a merged group. Its length must equal `length(sampleLabels)`. Use consecutive positive integers beginning with `1`, such as `c(1, 1, 2, 2)`. The number of unique indices must equal `length(sampleLabelsMerged)`. |
| `refSource` | No | Zero or one path to an existing genome-annotation file used by `annotateCluster()`. Supply it when cluster annotation will be performed and no annotation table has otherwise been added to the object. |
| `organismName` | No | Zero or one character string describing the organism. This optional metadata is not required for TSS clustering. |

### Accepted input types

The value of `inputFilesType` is case-sensitive and must be one of the
following:

| `inputFilesType` | Input represented | Required file arrangement |
|:--|:--|:--|
| `"bam"` | Single-end BAM alignments | One BAM file per sample |
| `"bamPairedEnd"` | Paired-end BAM alignments | One BAM file per sample |
| `"bed"` | BED-formatted TSS data | One BED file per sample |
| `"tss"` | CTSS/TSS position-and-signal data | One file per sample |
| `"BigWig"` | BigWig signal data | One BigWig file per sample |
| `"TSStable"` | A tabular TSS matrix containing multiple samples | Exactly one table containing all sample columns |

For example, four original samples can be merged into two experimental groups
using `sampleLabelsMerged = c("control", "treated")` and
`mergeIndex = c(1, 1, 2, 2)`. Index `1` corresponds to `"control"`, and index
`2` corresponds to `"treated"`; the ordering of `sampleLabelsMerged` must
therefore match the numeric group indices.

If replicate merging is not required, omit both `sampleLabelsMerged` and
`mergeIndex`. `TSSr()` will automatically retain each original sample as a
separate group. Do not supply only one of these two arguments.

## Importing TSS data

`getTSS()` reads the input file and stores the imported signal in the raw TSS
matrix. Displaying its dimensions provides a quick confirmation that the data
were imported before later analysis steps are run.

```{r import-tss, message=FALSE}
myTSSr <- getTSS(myTSSr)
dim(TSSmatrix(myTSSr, data = "raw"))
```

For formats other than `"TSStable"`, the import step may require additional
format-specific settings. See `?getTSS` before adapting the example to BAM or
other position-based files.

# TSS data processing

The `myTSSr` object contains chromosome I data and the upstream inputs
needed for each workflow stage. The chunks below recompute the complete
workflow and assign the returned object at every step.

```{r TSSprocessing, tidy=FALSE}
# Merge replicates
myTSSr <- mergeSamples(myTSSr)
# Normalization
myTSSr <- normalizeTSS(myTSSr)
# TSS filtering
rowsBeforeFiltering <- nrow(TSSmatrix(myTSSr, data = "processed"))
myTSSr <- filterTSS(myTSSr, method = "TPM", tpmLow = 2)
c(
    before = rowsBeforeFiltering,
    after = nrow(TSSmatrix(myTSSr, data = "processed"))
)
```

These functions populate the processed TSS matrix in the returned object. The
reported row counts show how many genomic TSS positions remain after the TPM
filter. Use `TSSmatrix()` to inspect this matrix through the public accessor:

```{r TSSprocessing-show}
# Access the processed TSS matrix without exposing internal storage
head(TSSmatrix(myTSSr, data = "processed"))
```

The first three columns identify the chromosome, genomic position, and strand;
the remaining columns contain processed signals for the merged sample groups.
This matrix is the input for TSS clustering.

# TSS clustering

Nearby TSSs can represent transcription initiation from the same core
promoter. `clusterTSS()` groups these positions into sample-specific tag
clusters using the peak-based `peakclu` method. Here, candidate peaks must be
at least 100 bp apart, a cluster can extend across gaps of at most 30 bp, TSSs
below 2% of the local peak are removed, and clusters with signal below 1 TPM
are discarded. `consensusCluster()` then relates tag clusters across sample
groups whose dominant TSSs are within 50 bp, providing common regions for
comparative analyses.

```{r TSSclustering, tidy=FALSE}
# TSS clustering
myTSSr <- clusterTSS(myTSSr,
    method = "peakclu", peakDistance = 100, extensionDistance = 30,
    localThreshold = 0.02, clusterThreshold = 1,
    useMultiCore = FALSE, numCores = NULL
)

# Aggregating consensus clusters
myTSSr <- consensusCluster(myTSSr, dis = 50, useMultiCore = FALSE)
```

The returned object now contains both `tagClusters` and `consensusClusters`.
Their accessors expose the sample-specific calls and the cross-sample regions,
respectively:

```{r TSSclustering-show}
# Tag clusters per sample
head(tagClusters(myTSSr, sample = "control"))

# Consensus clusters across samples
head(consensusClusters(myTSSr, sample = "control"))
```

Each table reports cluster coordinates, the dominant TSS, TSS signal, and
inter-quantile boundaries. Consensus clusters are used below so that promoter
shape, annotation, and between-group comparisons refer to the same genomic
regions. See `?clusterTSS` and `?consensusCluster` for parameter details.

# Core promoter shape

Core promoter shape describes how initiation signals are distributed within a
cluster. `shapeCluster()` can calculate either the Shape Index (SI) or Promoter
Shape Score (PSS); this example calculates PSS for the consensus clusters. PSS
combines the inter-quantile width with the distribution of TSS signals. Smaller
PSS values indicate sharper promoters, and a singleton has a score of zero.

```{r shapeCluster, tidy=FALSE}
# Calculating core promoter shape score
myTSSr <- shapeCluster(myTSSr,
    clusters = "consensusClusters", method = "PSS",
    useMultiCore = FALSE, numCores = NULL
)
```

The scores are stored in `clusterShape` for each merged sample group and can be
retrieved without accessing the object's internal slots:

```{r shapeCluster-show}
# Shape scores per cluster
head(clusterShape(myTSSr, sample = "control"))
```

These values can be used to compare promoter architectures across samples or
exported with `exportShapeTable()`. Only the most recently calculated shape
method is retained; see `?shapeCluster` for the SI alternative.

# Annotation of core promoters

Annotation connects consensus clusters to downstream genes so that subsequent
analyses can operate at gene level. The `refSource` supplied when `myTSSr` was
constructed points to the bundled GFF annotation. With the settings below, a
cluster is considered up to 1000 bp upstream of a gene start; overlapping
upstream features use the more restrictive 500-bp rule. Weak downstream
clusters below 2% of the strongest cluster in the promoter region are filtered
to reduce recapping and other low-signal events.

```{r annotateCluster, tidy=FALSE}
# Assign clusters to the annotated features
myTSSr <- annotateCluster(myTSSr, clusters = "consensusClusters",
    filterCluster = TRUE, filterClusterThreshold = 0.02,
    annotationType = "genes", upstream = 1000,
    upstreamOverlap = 500, downstream = 0)
```

The returned object separates clusters into assigned, unassigned, and, when
filtering is enabled, filtered tables. The following accessor displays clusters
that were assigned to genes in the control group:

```{r annotateCluster-show}
# Clusters assigned to genes
head(assignedClusters(myTSSr, sample = "control"))
```

The `gene` column records the assigned feature, while the cluster columns retain
the genomic position and signal information. Assigned clusters support the
gene-level differential expression and promoter-shift analyses below;
unassigned clusters are considered during enhancer identification. See
`?annotateCluster` when adapting promoter windows to another organism.

# Enhancer identification

Active enhancers can produce nearby, divergent transcription signals.
`callEnhancer()` searches the unassigned clusters for opposite-strand pairs no
more than 400 bp apart, applies a directionality filter, and excludes candidates
within 2000 bp of a gene's main annotated promoter. The result should therefore
be interpreted as a set of putative enhancers rather than independently
validated regulatory elements.

```{r callEnhancer, tidy=FALSE}
myTSSr <- callEnhancer(myTSSr, flanking = 400, dis2gene = 2000)
```

Candidate pairs and their strand-specific signals are stored in the `enhancers`
slot. Use the public accessor to inspect one sample group:

```{r callEnhancer-show}
# Putative enhancers per sample
head(enhancers(myTSSr, sample = "control"))
```

The output identifies the paired clusters, their dominant TSS positions and
signals, and the directionality score `D`. See `?callEnhancer` for the candidate
selection parameters and `exportEnhancerTable()` for saving the complete table.

# Differential expression analysis

After annotation, TSSr can aggregate raw TSS counts from assigned clusters by
gene and use DESeq2 to compare sample groups. `comparePairs` sets the order of
the comparison, here `control` versus `treat`, and `pval = 0.01` retains genes
with an adjusted p value below 0.01 in the significant-results table. DESeq2 is
a suggested dependency and must be installed to run this step. The bundled
example contains only a few annotated genes, so it uses `fitType = "mean"` to
avoid an unstable local dispersion-trend fit. The default remains
`fitType = "parametric"` for ordinary analyses.

```{r deGene, tidy=FALSE}
# Gene-level differential expression using DESeq2
myTSSr <- deGene(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01, useMultiCore = FALSE, numCores = NULL,
    fitType = "mean")
```

The returned object stores the gene-level count tables and both complete and
significant DESeq2 results. `DEtables()` selects a comparison and result set:

```{r deGene-show}
# Differential expression results
head(DEtables(
    myTSSr,
    comparison = "control_VS_treat",
    result = "significant"
))
```

Each row corresponds to one annotated gene and includes the DESeq2 effect-size
and significance statistics. Use `result = "all"` to retrieve the unfiltered
table, `plotDE()` to visualize it, or `exportDETable()` to save it.

# Core promoter shifts

Genes with multiple core promoters may use them at different proportions
between conditions. `shiftPromoter()` tests the annotated promoter profiles for
the requested `control`-versus-`treat` comparison. Sparse gene-level 2-by-2
tables can have small expected counts; in that case, the function emits one
summary warning for all affected tests and their approximate p values may be
unreliable. See `?shiftPromoter` for details.

```{r shiftPromoter, tidy=FALSE}
# Calculate core promoter shifts
myTSSr <- shiftPromoter(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01)
```

The test results are stored by comparison in the `PromoterShift` slot. The
accessor below retrieves the table for this pair:

```{r shiftPromoter-show}
# Promoter shift results
head(PromoterShift(myTSSr, comparison = "control_VS_treat"))
```

The table reports gene-level promoter-shift statistics and associated p values.
Because small expected counts weaken the chi-squared approximation, inspect the
underlying promoter signals before interpreting affected genes.

# Exporting results

TSSr can save processed TSS values and cluster tables for downstream analysis,
or write genome-browser tracks for visualization. The functions below write to
the current working directory. The vignette temporarily switches to
`tempdir()` so that rendering does not add result files to the package source.

```{r export}
oldwd <- setwd(tempdir())

# Export processed TSS values to ALL.samples.TSS.processed.txt
exportTSStable(myTSSr, data = "processed")

# Export cluster tables
exportClustersTable(myTSSr, data = "assigned")

# Export to BED format for genome browsers
exportClustersToBed(myTSSr, data = "consensusClusters")

# Export TSS to bedGraph
exportTSStoBedgraph(myTSSr, data = "processed")

setwd(oldwd)
```

This creates a tab-delimited processed TSS table, per-sample assigned-cluster
tables, BED files for assigned consensus clusters, and strand-specific bedGraph
tracks for the processed TSS signal. BED and bedGraph files can be loaded into
genome browsers such as UCSC Genome Browser or IGV. In an interactive analysis,
set the working directory to the desired output location before calling these
functions; see the individual export help pages for file-naming options.

# Precomputed example object

The package also includes `exampleTSSr`, a precomputed object for users who
want to inspect a populated `TSSr` object without rerunning the workflow. It is
not used as the input to the analysis above and does not replace `myTSSr`.

```{r precomputed-example}
data("exampleTSSr")
exampleTSSr
```

# Complete workflow in one pipeline

The step-by-step assignments above make intermediate results easy to inspect.
Once the settings have been established, the same analysis can be expressed
as one base R pipeline. The object is still named `myTSSr`; only the style of
the assignment changes. Replace the constructor values with the paths, labels,
genome, and annotation for your own experiment.

```{r complete-pipeline, eval=FALSE}
myTSSr <- TSSr(
    genomeName = "BSgenome.Scerevisiae.UCSC.sacCer3",
    inputFiles = exampleInput,
    inputFilesType = "TSStable",
    sampleLabels = c("SL01", "SL02", "SL03", "SL04"),
    sampleLabelsMerged = c("control", "treat"),
    mergeIndex = c(1, 1, 2, 2),
    refSource = exampleAnnotation
) |>
    getTSS() |>
    mergeSamples() |>
    normalizeTSS() |>
    filterTSS(method = "TPM", tpmLow = 2) |>
    clusterTSS(
        method = "peakclu",
        peakDistance = 100,
        extensionDistance = 30,
        localThreshold = 0.02,
        clusterThreshold = 1,
        useMultiCore = FALSE,
        numCores = NULL
    ) |>
    consensusCluster(
        dis = 50,
        useMultiCore = FALSE
    ) |>
    shapeCluster(
        clusters = "consensusClusters",
        method = "PSS",
        useMultiCore = FALSE,
        numCores = NULL
    ) |>
    annotateCluster(
        clusters = "consensusClusters",
        filterCluster = TRUE,
        filterClusterThreshold = 0.02,
        annotationType = "genes",
        upstream = 1000,
        upstreamOverlap = 500,
        downstream = 0
    ) |>
    callEnhancer(
        flanking = 400,
        dis2gene = 2000
    ) |>
    deGene(
        comparePairs = list(c("control", "treat")),
        pval = 0.01,
        useMultiCore = FALSE,
        numCores = NULL,
        fitType = "mean"
    ) |>
    shiftPromoter(
        comparePairs = list(c("control", "treat")),
        pval = 0.01
    )
```

The export functions are intentionally kept outside this pipeline because they
write files rather than add a new analysis stage to the returned object.

# Session info

Recording package and R versions makes the rendered analysis easier to
reproduce and diagnose:

```{r sessionInfo, tidy=FALSE}
sessionInfo()
```
