---
title: "TSSr Vignette"
author: "Zhaolian Lu, Keenan Berry, Zhenbin Hu, Yu Zhan, Tae-Hyuk Ahn, Zhenguo Lin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TSSr Vignette}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

TSSr is a comprehensive R/Bioconductor package for analyzing transcription
start site (TSS) sequencing data. It supports multiple input formats including
BAM, BED, BigWig, and TSS tables, and provides a complete workflow from TSS
identification through downstream analyses such as core promoter shape analysis,
gene annotation, differential expression, and promoter shifting.

## Installation

Install TSSr through Bioconductor with `BiocManager`:

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("TSSr")
```

And load `TSSr`:

```{r library, results='hide', message=FALSE}
library(TSSr)
```

# Citation

If you use TSSr, please cite the following article:

```{r citation, eval=TRUE}
citation("TSSr")
```

# Getting help

For general questions about the usage of TSSr, use the [official Bioconductor support forum](https://support.bioconductor.org) and tag your question "TSSr". We strive to answer questions as quickly as possible.

For technical questions, bug reports and suggestions for new features, we refer to the [TSSr github page](https://github.com/Linlab-slu/TSSr/issues).

# Quick start

```{r myTSSr-data, results="hide", tidy=FALSE}
# Load the example data
data("exampleTSSr")
myTSSr <- exampleTSSr
```

The bundled object is precomputed so that the main sections below can
demonstrate every downstream analysis with a sufficiently rich dataset. Its
source BAM paths are intentionally empty; `inputFilesType = "bam"` is retained
only as provenance, and its annotation is already stored in `refTable`.

## Starting from your own data

For a new analysis, `TSSr()` requires the genome name, input paths, input type,
and sample labels. Input paths must be existing regular files. `refSource` is
optional at construction, but when supplied it must also exist; annotation
later requires either `refSource` or a populated `refTable`.

The package includes a small four-sample TSS table and GFF file that exercise
the same import-to-annotation path used for user files:

```{r own-data-input}
exampleInput <- system.file(
    "extdata", "example-tss-table.tsv", package = "TSSr", mustWork = TRUE
)
exampleAnnotation <- system.file(
    "extdata", "example-annotation.gff3", package = "TSSr", mustWork = TRUE
)
importedTSSr <- TSSr(
    genomeName = "BSgenome.Scerevisiae.UCSC.sacCer3",
    inputFiles = exampleInput,
    inputFilesType = "TSStable",
    sampleLabels = c("SL01", "SL02", "SL03", "SL04"),
    sampleLabelsMerged = c("control", "treat"),
    mergeIndex = c(1, 1, 2, 2),
    refSource = exampleAnnotation
)
```

Importing is only the first stage. Before annotation, create tag clusters and
consensus clusters explicitly:

```{r own-data-workflow, message=FALSE}
importedTSSr <- getTSS(importedTSSr)
importedTSSr <- mergeSamples(importedTSSr)
importedTSSr <- normalizeTSS(importedTSSr)
importedTSSr <- filterTSS(importedTSSr, method = "TPM", tpmLow = 2)
importedTSSr <- clusterTSS(
    importedTSSr,
    method = "peakclu",
    clusterThreshold = 1,
    useMultiCore = FALSE
)
importedTSSr <- consensusCluster(importedTSSr, useMultiCore = FALSE)
importedTSSr <- annotateCluster(
    importedTSSr,
    clusters = "consensusClusters",
    filterCluster = TRUE
)
```

Both assigned and unassigned clusters are now available:

```{r own-data-results}
head(assignedClusters(importedTSSr, sample = "control"))
head(unassignedClusters(importedTSSr, sample = "control"))
```

# TSS data processing

The `myTSSr` object contains chromosome I data and the upstream inputs
needed for each workflow stage. The chunks below recompute the complete
workflow and assign the returned object at every step.

```{r TSSprocessing, tidy=FALSE}
# Merge replicates
myTSSr <- mergeSamples(myTSSr)
# Normalization
myTSSr <- normalizeTSS(myTSSr)
# TSS filtering
rowsBeforeFiltering <- nrow(TSSmatrix(myTSSr, data = "processed"))
myTSSr <- filterTSS(myTSSr, method = "TPM", tpmLow = 2)
c(
    before = rowsBeforeFiltering,
    after = nrow(TSSmatrix(myTSSr, data = "processed"))
)
```

These functions populate the processed TSS matrix in the returned object. The
reported row counts show how many genomic TSS positions remain after the TPM
filter. Use `TSSmatrix()` to inspect this matrix through the public accessor:

```{r TSSprocessing-show}
# Access the processed TSS matrix without exposing internal storage
head(TSSmatrix(myTSSr, data = "processed"))
```

The first three columns identify the chromosome, genomic position, and strand;
the remaining columns contain processed signals for the merged sample groups.
This matrix is the input for TSS clustering.

# TSS clustering

Nearby TSSs can represent transcription initiation from the same core
promoter. `clusterTSS()` groups these positions into sample-specific tag
clusters using the peak-based `peakclu` method. Here, candidate peaks must be
at least 100 bp apart, a cluster can extend across gaps of at most 30 bp, TSSs
below 2% of the local peak are removed, and clusters with signal below 1 TPM
are discarded. `consensusCluster()` then relates tag clusters across sample
groups whose dominant TSSs are within 50 bp, providing common regions for
comparative analyses.

```{r TSSclustering, tidy=FALSE}
# TSS clustering
myTSSr <- clusterTSS(myTSSr,
    method = "peakclu", peakDistance = 100, extensionDistance = 30,
    localThreshold = 0.02, clusterThreshold = 1,
    useMultiCore = FALSE, numCores = NULL
)

# Aggregating consensus clusters
myTSSr <- consensusCluster(myTSSr, dis = 50, useMultiCore = FALSE)
```

The returned object now contains both `tagClusters` and `consensusClusters`.
Their accessors expose the sample-specific calls and the cross-sample regions,
respectively:

```{r TSSclustering-show}
# Tag clusters per sample
head(tagClusters(myTSSr, sample = "control"))

# Consensus clusters across samples
head(consensusClusters(myTSSr, sample = "control"))
```

Each table reports cluster coordinates, the dominant TSS, TSS signal, and
inter-quantile boundaries. Consensus clusters are used below so that promoter
shape, annotation, and between-group comparisons refer to the same genomic
regions. See `?clusterTSS` and `?consensusCluster` for parameter details.

# Core promoter shape

Core promoter shape describes how initiation signals are distributed within a
cluster. `shapeCluster()` can calculate either the Shape Index (SI) or Promoter
Shape Score (PSS); this example calculates PSS for the consensus clusters. PSS
combines the inter-quantile width with the distribution of TSS signals. Smaller
PSS values indicate sharper promoters, and a singleton has a score of zero.

```{r shapeCluster, tidy=FALSE}
# Calculating core promoter shape score
myTSSr <- shapeCluster(myTSSr,
    clusters = "consensusClusters", method = "PSS",
    useMultiCore = FALSE, numCores = NULL
)
```

The scores are stored in `clusterShape` for each merged sample group and can be
retrieved without accessing the object's internal slots:

```{r shapeCluster-show}
# Shape scores per cluster
head(clusterShape(myTSSr, sample = "control"))
```

These values can be used to compare promoter architectures across samples or
exported with `exportShapeTable()`. Only the most recently calculated shape
method is retained; see `?shapeCluster` for the SI alternative.

# Annotation of core promoters

Annotation connects consensus clusters to downstream genes so that subsequent
analyses can operate at gene level. `myTSSr` already contains a 12-gene
reference table derived from the bundled GFF file. With the settings below, a
cluster is considered up to 1000 bp upstream of a gene start; overlapping
upstream features use the more restrictive 500-bp rule. Weak downstream
clusters below 2% of the strongest cluster in the promoter region are filtered
to reduce recapping and other low-signal events.

```{r annotateCluster, tidy=FALSE}
# Assign clusters to the annotated features
myTSSr <- annotateCluster(myTSSr, clusters = "consensusClusters",
    filterCluster = TRUE, filterClusterThreshold = 0.02,
    annotationType = "genes", upstream = 1000,
    upstreamOverlap = 500, downstream = 0)
```

The returned object separates clusters into assigned, unassigned, and, when
filtering is enabled, filtered tables. The following accessor displays clusters
that were assigned to genes in the control group:

```{r annotateCluster-show}
# Clusters assigned to genes
head(assignedClusters(myTSSr, sample = "control"))
```

The `gene` column records the assigned feature, while the cluster columns retain
the genomic position and signal information. Assigned clusters support the
gene-level differential expression and promoter-shift analyses below;
unassigned clusters are considered during enhancer identification. See
`?annotateCluster` when adapting promoter windows to another organism.

# Enhancer identification

Active enhancers can produce nearby, divergent transcription signals.
`callEnhancer()` searches the unassigned clusters for opposite-strand pairs no
more than 400 bp apart, applies a directionality filter, and excludes candidates
within 2000 bp of a gene's main annotated promoter. The result should therefore
be interpreted as a set of putative enhancers rather than independently
validated regulatory elements.

```{r callEnhancer, tidy=FALSE}
myTSSr <- callEnhancer(myTSSr, flanking = 400, dis2gene = 2000)
```

Candidate pairs and their strand-specific signals are stored in the `enhancers`
slot. Use the public accessor to inspect one sample group:

```{r callEnhancer-show}
# Putative enhancers per sample
head(enhancers(myTSSr, sample = "control"))
```

The output identifies the paired clusters, their dominant TSS positions and
signals, and the directionality score `D`. See `?callEnhancer` for the candidate
selection parameters and `exportEnhancerTable()` for saving the complete table.

# Differential expression analysis

After annotation, TSSr can aggregate raw TSS counts from assigned clusters by
gene and use DESeq2 to compare sample groups. `comparePairs` sets the order of
the comparison, here `control` versus `treat`, and `pval = 0.01` retains genes
with an adjusted p value below 0.01 in the significant-results table. DESeq2 is
a suggested dependency and must be installed to run this step.

```{r deGene, tidy=FALSE}
# Gene-level differential expression using DESeq2
myTSSr <- deGene(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01, useMultiCore = FALSE, numCores = NULL)
```

The returned object stores the gene-level count tables and both complete and
significant DESeq2 results. `DEtables()` selects a comparison and result set:

```{r deGene-show}
# Differential expression results
head(DEtables(
    myTSSr,
    comparison = "control_VS_treat",
    result = "significant"
))
```

Each row corresponds to one annotated gene and includes the DESeq2 effect-size
and significance statistics. Use `result = "all"` to retrieve the unfiltered
table, `plotDE()` to visualize it, or `exportDETable()` to save it.

# Core promoter shifts

Genes with multiple core promoters may use them at different proportions
between conditions. `shiftPromoter()` tests the annotated promoter profiles for
the requested `control`-versus-`treat` comparison. Sparse gene-level 2-by-2
tables can have small expected counts; in that case, the function emits one
summary warning for all affected tests and their approximate p values may be
unreliable. See `?shiftPromoter` for details.

```{r shiftPromoter, tidy=FALSE}
# Calculate core promoter shifts
myTSSr <- shiftPromoter(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01)
```

The test results are stored by comparison in the `PromoterShift` slot. The
accessor below retrieves the table for this pair:

```{r shiftPromoter-show}
# Promoter shift results
head(PromoterShift(myTSSr, comparison = "control_VS_treat"))
```

The table reports gene-level promoter-shift statistics and associated p values.
Because small expected counts weaken the chi-squared approximation, inspect the
underlying promoter signals before interpreting affected genes.

# Exporting results

TSSr can save processed TSS values and cluster tables for downstream analysis,
or write genome-browser tracks for visualization. The functions below write to
the current working directory. The vignette temporarily switches to
`tempdir()` so that rendering does not add result files to the package source.

```{r export}
oldwd <- setwd(tempdir())

# Export processed TSS values to ALL.samples.TSS.processed.txt
exportTSStable(myTSSr, data = "processed")

# Export cluster tables
exportClustersTable(myTSSr, data = "assigned")

# Export to BED format for genome browsers
exportClustersToBed(myTSSr, data = "consensusClusters")

# Export TSS to bedGraph
exportTSStoBedgraph(myTSSr, data = "processed")

setwd(oldwd)
```

This creates a tab-delimited processed TSS table, per-sample assigned-cluster
tables, BED files for assigned consensus clusters, and strand-specific bedGraph
tracks for the processed TSS signal. BED and bedGraph files can be loaded into
genome browsers such as UCSC Genome Browser or IGV. In an interactive analysis,
set the working directory to the desired output location before calling these
functions; see the individual export help pages for file-naming options.

# Session info

Recording package and R versions makes the rendered analysis easier to
reproduce and diagnose:

```{r sessionInfo, tidy=FALSE}
sessionInfo()
```
