---
title: "TSSr Vignette"
author: "Zhaolian Lu, Keenan Berry, Zhenbin Hu, Yu Zhan, Tae-Hyuk Ahn, Zhenguo Lin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TSSr Vignette}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

TSSr is a comprehensive R/Bioconductor package for analyzing transcription
start site (TSS) sequencing data. It supports multiple input formats including
BAM, BED, BigWig, and TSS tables, and provides a complete workflow from TSS
identification through downstream analyses such as core promoter shape analysis,
gene annotation, differential expression, and promoter shifting.

## Installation

You can install the development version directly from GitHub using `devtools`:

```{r github, eval=FALSE}
devtools::install_github("Linlab-slu/TSSr")
```

And load `TSSr`:

```{r library, results='hide', message=FALSE}
library(TSSr)
```

# Citation

If you use TSSr, please cite the following article:

```{r citation, eval=TRUE}
citation("TSSr")
```

# Getting help

For general questions about the usage of TSSr, use the [official Bioconductor support forum](https://support.bioconductor.org) and tag your question "TSSr". We strive to answer questions as quickly as possible.

For technical questions, bug reports and suggestions for new features, we refer to the [TSSr github page](https://github.com/Linlab-slu/TSSr/issues).

# Quick start

```{r exampleTSSr, results="hide", tidy=FALSE}
# Load the example data
data("exampleTSSr")
myTSSr <- exampleTSSr
```

Or create a new TSSr object using the constructor function. Note that the
input files do not need to exist at object creation time:

```{r newTSSr, tidy=FALSE}
# Create a TSSr object using the constructor (files do not need to exist)
newObj <- TSSr(
    genomeName = "BSgenome.Scerevisiae.UCSC.sacCer3",
    inputFiles = c("S01.sorted.bam", "S02.sorted.bam",
                    "S03.sorted.bam", "S04.sorted.bam"),
    inputFilesType = "bam",
    sampleLabels = c("SL01", "SL02", "SL03", "SL04"),
    sampleLabelsMerged = c("control", "treat"),
    mergeIndex = c(1, 1, 2, 2),
    refSource = "saccharomyces_cerevisiae.SGD.gff",
    organismName = "saccharomyces cerevisiae"
)
newObj
```

Then read in TSS data from the input files (requires actual files):

```{r getTSS, eval=FALSE}
# Get TSS (requires input files to exist)
getTSS(myTSSr)
```

# TSS data processing

> The `exampleTSSr` object distributed with the package already contains
> the results of every pipeline step below. Each computational chunk is
> shown as `eval=FALSE` to keep the vignette build time small on
> Bioconductor's Single Package Builder; the chunk that follows it
> displays the corresponding pre-computed result from `exampleTSSr` so
> the reader can see what each step produces.

```{r TSSprocessing, eval=FALSE, tidy=FALSE}
# Merge replicates
mergeSamples(myTSSr)
# Normalization
normalizeTSS(myTSSr)
# TSS filtering
filterTSS(myTSSr, method = "TPM", tpmLow = 0.1)
```

```{r TSSprocessing-show}
# Result available in the @TSSprocessedMatrix slot
head(slot(myTSSr, "TSSprocessedMatrix"))
```

# TSS clustering

```{r TSSclustering, eval=FALSE, tidy=FALSE}
# TSS clustering
clusterTSS(myTSSr,
    method = "peakclu", peakDistance = 100, extensionDistance = 30,
    localThreshold = 0.02, clusterThreshold = 1,
    useMultiCore = FALSE, numCores = NULL
)

# Aggregating consensus clusters
consensusCluster(myTSSr, dis = 50, useMultiCore = FALSE)
```

```{r TSSclustering-show}
# Tag clusters per sample (@tagClusters)
head(slot(myTSSr, "tagClusters")[["control"]])

# Consensus clusters across samples (@consensusClusters)
head(slot(myTSSr, "consensusClusters")[["control"]])
```

# Core promoter shape

```{r shapeCluster, eval=FALSE, tidy=FALSE}
# Calculating core promoter shape score
shapeCluster(myTSSr,
    clusters = "consensusClusters", method = "PSS",
    useMultiCore = FALSE, numCores = NULL
)
```

```{r shapeCluster-show}
# Shape scores per cluster (@clusterShape)
head(slot(myTSSr, "clusterShape")[["control"]])
```

# Annotation of core promoters

```{r annotateCluster, eval=FALSE, tidy=FALSE}
# Assign clusters to the annotated features
annotateCluster(myTSSr, clusters = "consensusClusters",
    filterCluster = TRUE, filterClusterThreshold = 0.02,
    annotationType = "genes", upstream = 1000,
    upstreamOverlap = 500, downstream = 0)
```

```{r annotateCluster-show}
# Clusters assigned to genes (@assignedClusters)
head(slot(myTSSr, "assignedClusters")[["control"]])
```

# Differential expression analysis

```{r deGene, eval=FALSE, tidy=FALSE}
# Gene-level differential expression using DESeq2
deGene(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01, useMultiCore = FALSE, numCores = NULL)
```

```{r deGene-show}
# Differential expression results (@DEtables)
head(slot(myTSSr, "DEtables")[["control_VS_treat"]])
```

# Core promoter shifts

```{r shiftPromoter, eval=FALSE, tidy=FALSE}
# Calculate core promoter shifts
shiftPromoter(myTSSr, comparePairs = list(c("control", "treat")),
    pval = 0.01)
```

```{r shiftPromoter-show}
# Promoter shift results (@PromoterShift)
head(slot(myTSSr, "PromoterShift")[["control_VS_treat"]])
```

# Exporting results

TSSr provides several export functions for saving results:

```{r export, eval=FALSE}
# Export TSS tables
exportTSStable(myTSSr, data = "processed")

# Export cluster tables
exportClustersTable(myTSSr, data = "assigned")

# Export to BED format for genome browsers
exportClustersToBed(myTSSr, data = "consensusClusters")

# Export TSS to bedGraph
exportTSStoBedgraph(myTSSr, data = "processed")
```

# Session info

```{r sessionInfo, tidy=FALSE}
sessionInfo()
```
